Ethereum: A Secure Decentralised Generalised Transaction Ledger

### 2014-04-06 Paper.tex
Complete execution model.

\begin{enumerate}
\item The transaction signature is valid;
\item the transaction nonce is valid (equivalent to the sender account's current nonce);
\item the gas limit is no smaller than the intrinsic gas, $g_0$, used by the transaction;
\item the sender account balance contains at least the cost, $c$, (in wei) required in up-front payment;
\end{enumerate}

@@

\begin{eqnarray}
g_0 & \equiv & \begin{cases}
  |T_\mathbf{d}| G_{data} + G_{call} & \quad \text{call} \\
  (|T_\mathbf{i}| + |T_\mathbf{b}|) G_{data} + G_{create} & \quad \text{create} \\
\end{cases}
\end{eqnarray}

@@

\begin{equation}
v_0 \equiv (T_g + g_0) T_p + T_v
\end{equation}

@@

Evaluating $\mathbb{S}_C$ from $\mathbb{S}_0$ depends on the transaction type; either contract creation or message call; we define the pairing of post-execution provisional state ($\mathbb{S}_P$) and remaining gas ($g'$):

\begin{equation}
(\mathbb{S}_P, g') \equiv \begin{cases}
\Lambda(\mathbb{S}_0, S(T), T_o, T_g, T_p, T_v, T_\mathbf{i}, T_\mathbf{b}) & \text{if} \quad T_r = 0 \\
\Theta_{0,1}(\mathbb{S}_0, S(T), T_o, T_r, T_g, T_p, T_v, T_\mathbf{d}) & \text{otherwise}
\end{cases}
\end{equation}

Note we use $\Theta_{0,1}$ to take only the first two components of the function's value; the third represents the message-call's output value (a byte array) and is unused in the context of transaction evaluation.

@@

\begin{eqnarray}
\mathbb{S}' & \equiv & \mathbb{S}_o \quad \text{except} \\
\mathbb{S}'[s]_{balance} & \equiv & \mathbb{S}_o[s]_{balance} + g' p
\end{eqnarray}

@@

We define the creation function formally as the function $\Lambda$, which evaluates from these values, together with the state $\mathbb{S}_C$ to the tuple containing the new state, additional database entries and remaining gas $(\mathbb{S}_P, g')$, as in \ref{ch:transactions}:

\begin{equation}
(\mathbb{S}_P, g') \equiv \Lambda(\mathbb{S}_C, s, o, g, p, v, \mathbf{i}, \mathbf{b})
\end{equation}

TODO: Move rest to own section after execution model.

TODO: Transaction structure layout.

@@

\begin{eqnarray}
A(s, o) & \equiv & a \quad \text{where:} \\
a' & \equiv & \mathcal{L}_{160}(\mathcal{S}(\mathcal{R}([s, o]))) \\
a & \equiv & \arg \min_x : x \geqslant a' \wedge \mathbb{S}_C[x] = \varnothing
\end{eqnarray}

where $\mathcal{S}$ is the SHA3 256-bit hash function, $\mathcal{R}$ is the RLP encoding function, $\mathcal{L}_y(X)$ evaluates to the rightmost $y$ bits of the binary data $X$ and $\mathbb{S}_C[x]$ is the address state of $x$ or $\varnothing$ if none exists.

@@

\begin{eqnarray}
(\mathbb{S}^{**}, g') & \equiv & \Xi(\mathbb{S}^*, g, I) \\
\mathbb{S}_P & \equiv & \begin{cases}
\mathbb{S}_C & \text{if} \quad \mathbb{S}^{**} = \varnothing \\
\mathbb{S}^{**} & \text{otherwise}
\end{cases} \\
I'_a & \equiv & a \\
I'_o & \equiv & s \\
I'_p & \equiv & p \\
I'_D & \equiv & [] \\
I'_s & \equiv & s \\
I'_v & \equiv & v \\
I'_C & \equiv & \mathbf{i}
\end{eqnarray}

$I'_D$ evaluates to the empty array. $I'_H$ and $I'_n$ have no special treatment and are determined from the blockchain.

@@

\begin{equation}
(\mathbb{S}_P, g', \mathbf{o}) \equiv \Theta(\mathbb{S}_C, s, o, r, g, p, v, \mathbf{d})
\end{equation}

In the case that the recipient contains code to be executed (i.e. that the account is a contract), then the contract's body code (identified as the fragment whose SHA3 hash is $\mathbb{S}[r]_{code}]$) is executed according to the execution model (see section \ref{ch:model}). Just as with contract creation, if the execution halts due to an exhausted gas supply, then no gas is refunded to the caller and the state is reverted to the point immediately prior to code execution.

\begin{eqnarray}
(\mathbb{S}^{**}, g') & \equiv & \Xi(\mathbb{S}^*, g, I) \\
\mathbb{S}' & \equiv & \begin{cases}
\mathbb{S}^* & \text{if} \quad \mathbb{S}^{**} = \varnothing \\
\mathbb{S}^{**} & \text{otherwise}
\end{cases} \\
I'_a & \equiv & a \\
I'_o & \equiv & o \\
I'_p & \equiv & p \\
I'_D & \equiv & \mathbf{d} \\
I'_s & \equiv & s \\
I'_v & \equiv & v \\
\text{Let} \mathcal{S}(I'_C) & = & \mathbb{S}[r]_{code}
\end{eqnarray}

We assert that the state database contains the entry $(\mathcal{S}(I'_C), I'_C)$.

@@

Additionally, when memory is accessed with {\small MSTORE}, {\small MLOAD}, {\small RETURN}, {\small SHA3}, {\small CALLDATA} or {\small CALL}, the memory should be enlarged to the smallest multiple of words such that all addressed bytes now fit in it. The memory size must always be a whole number of 32-byte words. Suppose that the highest previously accessed memory index is $i$, and the new index is $i'$; an additional $G_{m}(i, i')$ gas must be paid:

@@

\begin{itemize}
\item $I_a$, the address of the account which owns the code that is executing.
\item $I_o$, the sender address of the transaction that originated this execution.
\item $I_p$, the price of gas in the transaction that originated this execution.
\item $I_D$, the byte array that is the input data to this execution; if the execution agent is a transaction, this would be the transaction data.
\item $I_s$, the address of the account which caused the code to be executing; if the execution agent is a transaction, this would be the transaction sender.
\item $I_v$, the value passed to this account as part of the same procedure as execution; if the execution agent is a transaction, this would be the transaction value.
\item $I_C$, the byte array that is the machine code to be executed.
\item $I_H$, the block header of the previous block.
\item $I_n$, the block number of the current block.
\end{itemize}

@@

We must now define the $\Xi$ function. In most practical implementations this will be modelled as an iterative progression of the pair comprising the full system state, $\mathbb{S}$ and the machine state, $\mathbb{M}$. Formally, we define it recursively with a function $X$. This uses an iterator function $O$ (which defines the result of a single cycle of the state machine) together with functions $Z$ which determines if the present state is an exceptional halting state of the machine and $H$, specifying if the present state is a normal halting state of the machine:

\begin{eqnarray}
\Xi(\mathbb{S}, g, I) & \equiv & X(\mathbb{S}, \mathbb{M}, I) \\
\mathbb{M}_g & \equiv & g \\
\mathbb{M}_{pc} & \equiv & 0 \\
\mathbb{M}_M & \equiv & [0, 0, ...] \\
\mathbb{M}_i & \equiv & -1 \\
\mathbb{M}_T & \equiv & [] \\
X(\mathbb{S}, \mathbb{M}, I) & \equiv & \begin{cases}
(\mathbb{S}, \mathbb{M}, I) & \text{if} \quad Z(\mathbb{S}, \mathbb{M}, I)\\
O(\mathbb{S}, \mathbb{M}, I) & \text{if} \quad H(\mathbb{S}, \mathbb{M}, I)\\
X(O(\mathbb{S}, \mathbb{M}, I)) & \text{otherwise}\\
\end{cases}
\end{eqnarray}

@@

For the ease of reading, the instruction mnemonics, written in smallcaps (\eg {\small ADD}), should be interpreted as their numeric equivalents; the full table of instructions and their specifics is given in Appendix \ref{app:vm}.

@@

For the purposes of defining $H$ and $O$, we define $o$ as the current operation to be executed:

\begin{equation}
o \equiv \begin{cases} I_C[\mathbb{M}_{pc}] & \text{if} \quad \mathbb{M}_{pc} < \lVert I_C \rVert \\
\text{\small STOP} & \text{otherwise}
\end{cases}
\end{equation}

We also assume the fixed amounts of $\mathbf{\delta}$ and $\mathbf{\alpha}$, specifying the stack movement items removed and added, both subscriptable on the instruction and an instruction cost function $C$ evaluating to the full cost, in gas, of executing the given instruction. We also institute a function

For calculating the exceptional halting function $Z$, we assume a secondary, instruction-specific halting function $Y$, which identifies edge cases specific to particular instructions (such as the division by zero or initiating a message call with insufficient funds). Thus $Z$ is defined as:

\begin{equation}
\begin{array} {lll}
Z(\mathbb{S}, \mathbb{M}, I) & \equiv & Y(\mathbb{S}, \mathbb{M}, I) \\
& & \vee \quad \mathbb{M}_g < C(\mathbb{S}, \mathbb{M}, I) \\
& & \vee \quad \lVert\mathbb{M}_T\rVert < \mathbf{\delta}_o
\end{array}
\end{equation}

This states that the execution is in an exceptional halting state if there is an instruction-specific issue, if there is insufficient gas or if there are insufficient stack items.

@@

Stack items are added or removed from the left-most, lower-indexed portion of the series; all other items remain unchanged:

\begin{eqnarray}
O(\mathbb{S}, \mathbb{M}, I) & \equiv & (\mathbb{S}', \mathbb{M}', I) \\
\Delta & \equiv & \mathbf{\alpha}_o - \mathbf{\delta}_o \\
\lVert\mathbb{M}'_{T}\rVert & \equiv & \lVert\mathbb{M}_{T}\rVert + \Delta \\
\quad \forall i \in [\mathbf{\alpha}_o, \lVert\mathbb{M}'_T\rVert): \mathbb{M}'_{T_i} & \equiv & \mathbb{M}_{T_{i+\Delta}}
\end{eqnarray}

The gas is reduced by the instruction's gas cost and for most instructions, the program counter increments on each cycle, for the three exceptions, we assume a function $J$ which evaluates to the according value:

\begin{eqnarray}
\mathbb{M}'_{g} \equiv \mathbb{M}_{g} - C(\mathbb{S}, \mathbb{M}, I) \\
\mathbb{M}'_{pc} \equiv \begin{cases}
J(\mathbb{M}) & \text{if} \quad o \in \{ \text{\small JUMP, JUMPI} \} \\
\mathbb{M}_{pc} + p & \text{if} \quad o \in [\text{\small PUSH1}, \text{\small PUSH32}] \\
\mathbb{M}_{pc} + 1 & \text{otherwise}
\end{cases}
\end{eqnarray}

where $p$ is the byte size of the push instruction, defined as:

\begin{equation}
p \equiv o - \text{\small PUSH1} + 2
\end{equation}

In general, we assume the memory and system state don't change:

\begin{eqnarray}
\mathbb{M}'_M & \equiv & \mathbb{M}_M \\
\mathbb{M}'_i & \equiv & \mathbb{M}_i \\
\mathbb{S}' & \equiv & \mathbb{S}
\end{eqnarray}

However, instructions do typically alter one or several components of these values. Altered components listed by instruction are noted in Appendix \ref{app:vm}.

@@

\begin{tabular*}{\columnwidth}[h]{rlrrl}
\toprule
\multicolumn{5}{c}{\textbf{30s: Environmental Information}} \vspace{5pt} \\
\textbf{Value} & \textbf{Mnemonic} & $\delta$ & $\alpha$ & \textbf{Description} \vspace{5pt} \\
0x30 & {\small ADDRESS} & 0 & 1 & Get address of currently executing account. \\
&&&& $\mathbb{M}'_T[0] \equiv I_a$ \\
\midrule
0x31 & {\small BALANCE} & 0 & 1 & Get balance of currently executing account. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{S}[I_a]_{balance}$ \\
\midrule
0x32 & {\small ORIGIN} & 0 & 1 & Get execution origination address. \\
&&&& $\mathbb{M}'_T[0] \equiv I_o$ \\
&&&& This is the sender of original transaction; it's never a contract. \\
\midrule
0x33 & {\small CALLER} & 0 & 1 & Get caller address. \\
&&&& $\mathbb{M}'_T[0] \equiv I_c$ \\
&&&& This is the address of the account that is directly responsible for this execution. \\
\midrule
0x34 & {\small CALLVALUE} & 0 & 1 & Get deposited value by the instruction/transaction responsible for this execution. \\
&&&& $\mathbb{M}'_T[0] \equiv I_v$ \\
\midrule
0x35 & {\small CALLDATALOAD} & 1 & 1 & Get input data of current environment. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathcal{I}_D[ \mathbb{M}_S[0] \dots (\mathbb{M}_S[0] + 31) ]$ \\
&&&& This pertains to the input data passed with the message call instruction or transaction. \\
\midrule
0x36 & {\small CALLDATASIZE} & 0 & 1 & Get size of input data in current environment. \\
&&&& $\mathbb{M}'_T[0] \equiv \lVert \mathcal{I}_D \rVert$ \\
&&&& This pertains to the input data passed with the message call instruction or transaction. \\
\midrule
0x37 & {\small GASPRICE} & 0 & 1 & Get price of gas in current environment. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathcal{I}_p$ \\
&&&& This is gas price specified by the originating transaction.\\
\bottomrule
\end{tabular*}

@@

\begin{tabular*}{\columnwidth}[h]{rlrrl}
\toprule
\multicolumn{5}{c}{\textbf{50s: Stack, Memory, Storage and Flow Operations}} \vspace{5pt} \\
\textbf{Value} & \textbf{Mnemonic} & $\delta$ & $\alpha$ & \textbf{Description} \vspace{5pt} \\
0x50 & {\small POP} & 1 & 0 & Remove item from stack. \\
\midrule
0x51 & {\small DUP} & 1 & 2 & Duplicate stack item. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_T[0]$ \\
\midrule
0x52 & {\small SWAP} & 2 & 2 & Exchange stack items. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_T[1]$ \\
&&&& $\mathbb{M}'_T[1] \equiv \mathbb{M}_T[0]$ \\
\midrule
0x53 & {\small MLOAD} & 1 & 1 & Load word from memory. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_M[\mathbb{M}_T[0] \dots (\mathbb{M}_T[0] + 31) ]$ \\
\midrule
0x54 & {\small MSTORE} & 2 & 0 & Save word to memory. \\
&&&& $\mathbb{M}'_M[ \mathbb{M}_T[0] \dots (\mathbb{M}_T[0] + 31) ] \equiv \mathbb{M}_T[1]$ \\
\midrule
0x55 & {\small MSTORE8} & 2 & 0 & Save byte to memory. \\
&&&& $\mathbb{M}'_M[ \mathbb{M}_T[0] ] \equiv (\mathbb{M}_T[1] \mod 256) $ \\
\midrule
0x56 & {\small SLOAD} & 1 & 1 & Load word from storage. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_S[\mathbb{M}_T[0]]$ \\
\midrule
0x57 & {\small SSTORE} & 2 & 0 & Save word to storage. \\
&&&& $\mathbb{M}'_S[ \mathbb{M}_T[0] ] \equiv \mathbb{M}_T[1] $ \\
\midrule
0x58 & {\small JUMP} & 1 & 0 & Alter the program counter. \\
&&&& $J(\mathbb{M}) \equiv \mathbb{M}_T[0] $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$ \\
\midrule
0x59 & {\small JUMPI} & 2 & 0 & Conditionally alter the program counter. \\
&&&& $J(\mathbb{M}) \equiv \begin{cases} \mathbb{M}_T[0] & \text{if} \quad \mathbb{M}_T[0] = 0 \\ \mathbb{M}_{pc} + 1 & \text{otherwise} \end{cases} $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$ \\
\midrule
0x5a & {\small PC} & 0 & 1 & Get the program counter. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_{pc}$ \\
\midrule
0x5b & {\small MSIZE} & 0 & 1 & Get the size of active memory. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_{i}$ \\
\midrule
0x5c & {\small GAS} & 0 & 1 & Get the amount of available gas. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_{g}$ \\
\bottomrule
\end{tabular*}

>>>>

\begin{enumerate}
\item The transaction signature is valid;
\item the transaction nonce is valid (equivalent to the sender account's current nonce);
\item the gas limit is no smaller than the intrinsic gas, $g_0$, used by the transaction;
\item the sender account balance contains at least the cost, $v_0$, (in wei) required in up-front payment;
\end{enumerate}

@@

\begin{eqnarray}
g_0 & \equiv & \begin{cases}
  (|T_\mathbf{i}| + |T_\mathbf{b}|) G_{data} + G_{create} & \text{if} \quad T_r = 0 \\
  |T_\mathbf{d}| G_{data} + G_{call} & \text{otherwise} \\
\end{cases}
\end{eqnarray}

@@

\begin{equation}
v_0 \equiv T_g T_p + T_v
\end{equation}

the validity is determined as:

\begin{eqnarray}
S(T) & \neq & \varnothing \\ \wedge \quad \mathbb{S}[S(T)] & \neq & \varnothing \\ \wedge \quad T_n & = & \mathbb{S}[S(T)]_{nonce} \\ \wedge \quad g_0 & \leq & T_g \\ \wedge \quad v_0 & \leq & \mathbb{S}[S(T)]_{balance}
\end{eqnarray}

@@

Evaluating $\mathbb{S}_0$ from $\mathbb{S}_0$ depends on the transaction type; either contract creation or message call; we define the pairing of post-execution provisional state ($\mathbb{S}_P$) and remaining gas ($g'$):

\begin{equation}
(\mathbb{S}_P, g') \equiv \begin{cases}
\Lambda(\mathbb{S}_0, S(T), T_o, g, T_p, T_v, T_\mathbf{i}, T_\mathbf{b}) & \text{if} \quad T_r = 0 \\
\Theta^*(\mathbb{S}_0, S(T), T_o, T_r, g, T_p, T_v, T_\mathbf{d}) & \text{otherwise}
\end{cases}
\end{equation}

where

\begin{equation}
g \equiv T_g - g_0
\end{equation}

Note we use $\Theta^*$ to take only the first two components of the function's value; the third represents the message-call's output value (a byte array) and is unused in the context of transaction evaluation.

@@

\begin{eqnarray}
\mathbb{S}' & \equiv & \mathbb{S}_o \quad \text{except} \\
\mathbb{S}'[s]_{balance} & \equiv & \mathbb{S}_o[s]_{balance} + g' T_p
\end{eqnarray}

@@

We define the creation function formally as the function $\Lambda$, which evaluates from these values, together with the state $\mathbb{S}_0$ to the tuple containing the new state, additional database entries and remaining gas $(\mathbb{S}_P, g')$, as in \ref{ch:transactions}:

\begin{equation}
(\mathbb{S}_P, g') \equiv \Lambda(\mathbb{S}_0, s, o, g, p, v, \mathbf{i}, \mathbf{b})
\end{equation}

@@

\begin{eqnarray}
A(s, o) & \equiv & a \quad \text{where:} \\
a' & \equiv & \mathcal{B}_{96..255}(\mathcal{S}(\mathcal{R}([s, o]))) \\
a & \equiv & \arg \min_x : x \geqslant a' \wedge \mathbb{S}_0[x] = \varnothing
\end{eqnarray}

where $\mathcal{S}$ is the SHA3 256-bit hash function, $\mathcal{R}$ is the RLP encoding function, $\mathcal{B}_{a..b}(X)$ evaluates to binary value containing the bits of indices in the range $[a, b]$ of the binary data $X$ and $\mathbb{S}_0[x]$ is the address state of $x$ or $\varnothing$ if none exists.

@@

\begin{eqnarray}
(\mathbb{S}^{**}, g') & \equiv & \Xi(\mathbb{S}^*, g, I) \\
\mathbb{S}_P & \equiv & \begin{cases}
\mathbb{S}_0 & \text{if} \quad \mathbb{S}^{**} = \varnothing \\
\mathbb{S}^{**} & \text{otherwise}
\end{cases} \\
I_a & \equiv & a \\
I_o & \equiv & s \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & [] \\
I_s & \equiv & s \\
I_v & \equiv & v \\
I_\mathbf{b} & \equiv & \mathbf{i}
\end{eqnarray}

$I_\mathbf{d}$ evaluates to the empty array. $I_H$ and $I_n$ have no special treatment and are determined from the blockchain.

@@

\begin{equation}
(\mathbb{S}_P, g', \mathbf{o}) \equiv \Theta(\mathbb{S}, s, o, r, g, p, v, \mathbf{d})
\end{equation}

We define $\mathbb{S}_1$, the rollback state as the original state but with the value transferred to the recipient:

\begin{eqnarray}
\mathbb{S}_1 & \equiv & \mathbb{S} \quad \text{except:} \\
\mathbb{S}_1[r]_{balance} & \equiv & \mathbb{S}_1[r]_{balance} + v
\end{eqnarray}

If the recipient account contains no code to be executed (i.e. it's not a contract), then we define the final state as being equivalent to the rollback state. In the case that the recipient contains code to be executed (i.e. that the account is a contract), then the contract's body code (identified as the fragment whose SHA3 hash is $\mathbb{S}[r]_{code}]$) is executed according to the execution model (see section \ref{ch:model}). Just as with contract creation, if the execution halts due to an exhausted gas supply, then no gas is refunded to the caller and the state is reverted to the point immediately prior to code execution.

\begin{eqnarray}
\mathbb{S}' & \equiv & \begin{cases}
\mathbb{S}_1 & \text{if} \quad \mathbb{S}_1[r]_{code} = \varnothing \\
\mathbb{S}_1 & \text{if} \quad \mathbb{S}^{**} = \varnothing \\
\mathbb{S}^{**} & \text{otherwise}
\end{cases} \\
(\mathbb{S}^{**}, g', \mathbf{o}) & \equiv & \Xi(\mathbb{S}_1, g, I) \\
I_a & \equiv & a \\
I_o & \equiv & o \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & \mathbf{d} \\
I_s & \equiv & s \\
I_v & \equiv & v \\
\text{Let} \quad \mathcal{S}(I_\mathbf{b}) & = & \mathbb{S}[r]_{code}
\end{eqnarray}

We assert that the state database contains the entry $(\mathcal{S}(I_\mathbf{b}), I_\mathbf{b})$.

@@

Additionally, when memory is accessed with {\small MSTORE}, {\small MLOAD}, {\small RETURN}, {\small SHA3}, {\small CALLDATA} or {\small CALL}, the memory should be enlarged to the smallest multiple of words such that all addressed bytes now fit in it. The memory size must always be a whole number of 32-byte words. Suppose that the highest previously accessed memory word index is $i$, and the new word index is $i'$; an additional $G_{m}(i, i')$ gas must be paid:

@@

\begin{itemize}
\item $I_a$, the address of the account which owns the code that is executing.
\item $I_o$, the sender address of the transaction that originated this execution.
\item $I_p$, the price of gas in the transaction that originated this execution.
\item $I_\mathbf{d}$, the byte array that is the input data to this execution; if the execution agent is a transaction, this would be the transaction data.
\item $I_s$, the address of the account which caused the code to be executing; if the execution agent is a transaction, this would be the transaction sender.
\item $I_v$, the value passed to this account as part of the same procedure as execution; if the execution agent is a transaction, this would be the transaction value.
\item $I_\mathbf{b}$, the byte array that is the machine code to be executed.
\item $I_H$, the block header of the previous block.
\item $I_n$, the block number of the current block.
\end{itemize}

@@

We must now define the $\Xi$ function. In most practical implementations this will be modelled as an iterative progression of the pair comprising the full system state, $\mathbb{S}$ and the machine state, $\mathbb{M}$. Formally, we define it recursively with a function $X$. This uses an iterator function $O$ (which defines the result of a single cycle of the state machine) together with functions $Z$ which determines if the present state is an exceptional halting state of the machine and $H$, specifying the output data of the instruction if and only if the present state is a normal halting state of the machine. We consider the empty sequence (denoted $[]$ to be non-equal to the empty set, denoted $\varnothing$).

\begin{eqnarray}
\Xi(\mathbb{S}, g, I) & \equiv & X(\mathbb{S}, \mathbb{M}, I) \\
\mathbb{M}_g & \equiv & g \\
\mathbb{M}_{pc} & \equiv & 0 \\
\mathbb{M}_M & \equiv & [0, 0, ...] \\
\mathbb{M}_i & \equiv & -1 \\
\mathbb{M}_T & \equiv & []
\end{eqnarray}
\begin{equation}
X(\mathbb{S}, \mathbb{M}, I) \equiv \begin{cases}
(\mathbb{S}, \mathbb{M}, I, []) & \text{if} \quad Z(\mathbb{S}, \mathbb{M}, I)\\
(O(\mathbb{S}, \mathbb{M}, I), \mathbf{o}) & \text{if} \quad \mathbf{o} \neq \varnothing\\
X(O(\mathbb{S}, \mathbb{M}, I)) & \text{otherwise}\\
\end{cases}
\end{equation}

where
\begin{equation}
\mathbf{o} \equiv H(\mathbb{S}, \mathbb{M}, I)
\end{equation}

@@

For the ease of reading, the instruction mnemonics, written in smallcaps (\eg \space {\small ADD}), should be interpreted as their numeric equivalents; the full table of instructions and their specifics is given in Appendix \ref{app:vm}.

For the purposes of defining $Z$, $H$ and $O$, we define $w$ as the current operation to be executed:

\begin{equation}
w \equiv \begin{cases} I_\mathbf{b}[\mathbb{M}_{pc}] & \text{if} \quad \mathbb{M}_{pc} < \lVert I_\mathbf{b} \rVert \\
\text{\small STOP} & \text{otherwise}
\end{cases}
\end{equation}

We also assume the fixed amounts of $\mathbf{\delta}$ and $\mathbf{\alpha}$, specifying the stack movement items removed and added, both subscriptable on the instruction and an instruction cost function $C$ evaluating to the full cost, in gas, of executing the given instruction.

@@

The exceptional halting function $Z$ is defined as:
\begin{equation}
Z(\mathbb{S}, \mathbb{M}, I) \equiv \mathbb{M}_g < C(\mathbb{S}, \mathbb{M}, I) \vee \lVert\mathbb{M}_T\rVert < \mathbf{\delta}_w
\end{equation}

This states that the execution is in an exceptional halting state if there is insufficient gas or if there are insufficient stack items. The astute read will realise that this implies that no instruction can, through its execution, cause an exceptional halt.

\subsubsection{Normal Halting}

We define the normal halting function $H$:
\begin{equation}
H(\mathbb{S}, \mathbb{M}, I) \equiv \begin{cases}
H_{\text{\tiny RETURN}}(\mathbb{M}) & \text{if} \quad w = \text{\small RETURN} \\
[] & \text{if} \quad w \in \{ \text{\small STOP}, \text{\small SUICIDE} \} \\
\varnothing & \text{otherwise}
\end{cases}
\end{equation}

The data-returning halt operation, \text{\small RETURN}, has a special function $H_{\text{\tiny RETURN}}$, defined in Appendix \ref{app:vm}.

@@

Stack items are added or removed from the left-most, lower-indexed portion of the series; all other items remain unchanged:
\begin{eqnarray}
O(\mathbb{S}, \mathbb{M}, I) & \equiv & (\mathbb{S}', \mathbb{M}', I) \\
\Delta & \equiv & \mathbf{\alpha}_w - \mathbf{\delta}_w \\
\lVert\mathbb{M}'_{T}\rVert & \equiv & \lVert\mathbb{M}_{T}\rVert + \Delta \\
\quad \forall x \in [\mathbf{\alpha}_w, \lVert\mathbb{M}'_T\rVert): \mathbb{M}'_{T_x} & \equiv & \mathbb{M}_{T_{x+\Delta}}
\end{eqnarray}

The gas is reduced by the instruction's gas cost and for most instructions, the program counter increments on each cycle, for the three exceptions, we assume a function $J$ which evaluates to the according value:
\begin{eqnarray}
\mathbb{M}'_{g} & \equiv & \mathbb{M}_{g} - C(\mathbb{S}, \mathbb{M}, I) \\
\mathbb{M}'_{pc} & \equiv & \begin{cases}
J_o(\mathbb{M}) & \text{if} \quad w \in \{ \text{\small JUMP, JUMPI} \} \\
\mathbb{M}_{pc} + p & \text{if} \quad w \in [\text{\small PUSH1}, \text{\small PUSH32}] \\
\mathbb{M}_{pc} + 1 & \text{otherwise}
\end{cases}
\end{eqnarray}

where $p$ is the byte size of the push instruction, defined as:
\begin{equation}
p \equiv w - \text{\small PUSH1} + 2
\end{equation}

In general, we assume the memory and system state don't change:
\begin{eqnarray}
\mathbb{M}'_M & \equiv & \mathbb{M}_M \\
\mathbb{M}'_i & \equiv & \mathbb{M}_i \\
\mathbb{S}' & \equiv & \mathbb{S}
\end{eqnarray}

However, instructions do typically alter one or several components of these values. Altered components listed by instruction are noted in Appendix \ref{app:vm}, alongside values for $\alpha$ and $\delta$.

@@

\begin{tabular*}{\columnwidth}[h]{rlrrl}
\toprule
\multicolumn{5}{c}{\textbf{30s: Environmental Information}} \vspace{5pt} \\
\textbf{Value} & \textbf{Mnemonic} & $\delta$ & $\alpha$ & \textbf{Description} \vspace{5pt} \\
0x30 & {\small ADDRESS} & 0 & 1 & Get address of currently executing account. \\
&&&& $\mathbb{M}'_T[0] \equiv I_a$ \\
\midrule
0x31 & {\small BALANCE} & 0 & 1 & Get balance of currently executing account. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{S}[I_a]_{balance}$ \\
\midrule
0x32 & {\small ORIGIN} & 0 & 1 & Get execution origination address. \\
&&&& $\mathbb{M}'_T[0] \equiv I_o$ \\
&&&& This is the sender of original transaction; it's never a contract. \\
\midrule
0x33 & {\small CALLER} & 0 & 1 & Get caller address. \\
&&&& $\mathbb{M}'_T[0] \equiv I_c$ \\
&&&& This is the address of the account that is directly responsible for this execution. \\
\midrule
0x34 & {\small CALLVALUE} & 0 & 1 & Get deposited value by the instruction/transaction responsible for this execution. \\
&&&& $\mathbb{M}'_T[0] \equiv I_v$ \\
\midrule
0x35 & {\small CALLDATALOAD} & 1 & 1 & Get input data of current environment. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathcal{I}_\mathbf{d}[ \mathbb{M}_S[0] \dots (\mathbb{M}_S[0] + 31) ]$ \\
&&&& This pertains to the input data passed with the message call instruction or transaction. \\
\midrule
0x36 & {\small CALLDATASIZE} & 0 & 1 & Get size of input data in current environment. \\
&&&& $\mathbb{M}'_T[0] \equiv \lVert \mathcal{I}_\mathbf{d} \rVert$ \\
&&&& This pertains to the input data passed with the message call instruction or transaction. \\
\midrule
0x37 & {\small GASPRICE} & 0 & 1 & Get price of gas in current environment. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathcal{I}_p$ \\
&&&& This is gas price specified by the originating transaction.\\
\bottomrule
\end{tabular*}

@@

\begin{tabular*}{\columnwidth}[h]{rlrrl}
\toprule
\multicolumn{5}{c}{\textbf{50s: Stack, Memory, Storage and Flow Operations}} \vspace{5pt} \\
\textbf{Value} & \textbf{Mnemonic} & $\delta$ & $\alpha$ & \textbf{Description} \vspace{5pt} \\
0x50 & {\small POP} & 1 & 0 & Remove item from stack. \\
\midrule
0x51 & {\small DUP} & 1 & 2 & Duplicate stack item. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_T[0]$ \\
\midrule
0x52 & {\small SWAP} & 2 & 2 & Exchange stack items. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_T[1]$ \\
&&&& $\mathbb{M}'_T[1] \equiv \mathbb{M}_T[0]$ \\
\midrule
0x53 & {\small MLOAD} & 1 & 1 & Load word from memory. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_M[\mathbb{M}_T[0] \dots (\mathbb{M}_T[0] + 31) ]$ \\
\midrule
0x54 & {\small MSTORE} & 2 & 0 & Save word to memory. \\
&&&& $\mathbb{M}'_M[ \mathbb{M}_T[0] \dots (\mathbb{M}_T[0] + 31) ] \equiv \mathbb{M}_T[1]$ \\
\midrule
0x55 & {\small MSTORE8} & 2 & 0 & Save byte to memory. \\
&&&& $\mathbb{M}'_M[ \mathbb{M}_T[0] ] \equiv (\mathbb{M}_T[1] \mod 256) $ \\
\midrule
0x56 & {\small SLOAD} & 1 & 1 & Load word from storage. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_S[\mathbb{M}_T[0]]$ \\
\midrule
0x57 & {\small SSTORE} & 2 & 0 & Save word to storage. \\
&&&& $\mathbb{M}'_S[ \mathbb{M}_T[0] ] \equiv \mathbb{M}_T[1] $ \\
\midrule
0x58 & {\small JUMP} & 1 & 0 & Alter the program counter. \\
&&&& $J_{\text{\tiny JUMP}}(\mathbb{M}) \equiv \mathbb{M}_T[0] $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$. See section \ref{ch:model}. \\
\midrule
0x59 & {\small JUMPI} & 2 & 0 & Conditionally alter the program counter. \\
&&&& $J_{\text{\tiny JUMPI}}(\mathbb{M}) \equiv \begin{cases} \mathbb{M}_T[0] & \text{if} \quad \mathbb{M}_T[0] = 0 \\ \mathbb{M}_{pc} + 1 & \text{otherwise} \end{cases} $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$. See section \ref{ch:model}. \\
\midrule
0x5a & {\small PC} & 0 & 1 & Get the program counter. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_{pc}$ \\
\midrule
0x5b & {\small MSIZE} & 0 & 1 & Get the size of active memory. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_{i}$ \\
\midrule
0x5c & {\small GAS} & 0 & 1 & Get the amount of available gas. \\
&&&& $\mathbb{M}'_T[0] \equiv \mathbb{M}_{g}$ \\
\bottomrule
\end{tabular*}
